---
title: "Random Forest Classification - Lung Cancer Prediction"
author: "Noelia Caba Martin"
date: "2026-03-18"
output: html_document
---

## Introduction

Lung cancer is one of the leading causes of cancer-related mortality 
worldwide. Early detection is critical for improving patient outcomes. 
Machine learning models trained on clinical and symptomatic data can 
serve as screening tools to identify high-risk patients.

In this analysis, we build a **Random Forest classification model** 
to predict lung cancer status from 15 clinical and demographic variables 
collected through patient surveys.

**Research question:** Can we accurately predict lung cancer presence 
from clinical symptoms and demographic variables?

The main objectives are:

- Perform exploratory data analysis and data cleaning
- Normalize and scale the data
- Train a Random Forest classification model
- Evaluate model performance using standard classification metrics
- Identify the most important predictors of lung cancer

## Setup
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```
```{r libraries}
library(tidyverse)
library(ggplot2)
library(randomForest)
library(caret)
library(pROC)
library(vip)
library(corrplot)
```

## Data Loading and Exploration
```{r load-data}
# Load dataset
lung <- read.csv("lung_cancer.csv")

# Convert to dataframe
lung <- data.frame(lung)

# Initial exploration
dim(lung)
str(lung)
head(lung)
```

The dataset contains **309 patients** and **16 variables** including 
demographic information, symptoms, and lifestyle factors. The target 
variable `LUNG_CANCER` is binary (YES/NO).

Key observations from the initial exploration:

- `GENDER` is coded as character (M/F) and requires conversion
- Most symptom variables are coded as integers (1 = No, 2 = Yes)
- `LUNG_CANCER` is coded as character (YES/NO) and requires conversion 
to factor for classification

## Data Cleaning and Preprocessing

Before training the model, we need to:

1. **Convert categorical variables** to appropriate types
2. **Check for missing values**
3. **Check class balance** of the target variable
4. **Normalize and scale** continuous variables
5. **Split** into training (80%) and test (20%) sets

```{r preprocessing}
# Check missing values
sum(is.na(lung))

# Convert GENDER to numeric (M=1, F=0)
lung$GENDER <- ifelse(lung$GENDER == "M", 1, 0)

# Convert LUNG_CANCER to factor
lung <- lung %>%
  mutate(across(where(is.integer), as.numeric))

# Convert LUNG_CANCER to factor
lung$LUNG_CANCER <- as.factor(lung$LUNG_CANCER)

# Check class balance
table(lung$LUNG_CANCER)
prop.table(table(lung$LUNG_CANCER))

# Normalize and scale AGE
lung$AGE <- scale(lung$AGE)

lung <- lung %>%
  mutate(across(where(is.integer), as.numeric))

# Split into training (80%) and test (20%) sets
set.seed(123)
train_index <- createDataPartition(lung$LUNG_CANCER,
                                   p = 0.8,
                                   list = FALSE)
train_data <- lung[train_index, ]
test_data <- lung[-train_index, ]

cat("Training set:", nrow(train_data), "observations\n")
cat("Test set:", nrow(test_data), "observations\n")
```

The preprocessing steps confirmed:

- **No missing values** in the dataset ✓
- **Class distribution:** 87.4% of patients have lung cancer (YES) 
and 12.6% do not (NO)
- The dataset shows class imbalance with the majority class being 
lung cancer positive
- **AGE** was normalized using z-score standardization (mean = 0, SD = 1)
- **GENDER** was converted to binary (M = 1, F = 0)
- The data was split into **248 training** and **61 test** observations

## Exploratory Data Analysis

We perform a brief exploratory analysis to visualize the distribution 
of key variables and understand the relationship between symptoms and 
lung cancer status.

```{r eda}
# Lung cancer distribution
ggplot(lung, aes(x = LUNG_CANCER, fill = LUNG_CANCER)) +
  geom_bar(alpha = 0.8, width = 0.5) +
  geom_text(stat = "count", 
            aes(label = after_stat(count)), 
            vjust = -0.5, size = 5, fontface = "bold") +
  scale_fill_manual(values = c("NO" = "#2E9FDF", "YES" = "#E64B35")) +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(title = "Lung Cancer Distribution",
       x = "Lung Cancer Status", y = "Count")

# Age distribution by lung cancer status
ggplot(lung, aes(x = LUNG_CANCER, y = AGE, fill = LUNG_CANCER)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = c("NO" = "#2E9FDF", "YES" = "#E64B35")) +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(title = "Age Distribution by Lung Cancer Status",
       x = "Lung Cancer Status", y = "Age (scaled)")

# Symptom prevalence by lung cancer status
lung %>%
  select(LUNG_CANCER, SMOKING, ANXIETY, CHRONIC.DISEASE, 
         FATIGUE, ALLERGY, WHEEZING, COUGHING) %>%
  pivot_longer(-LUNG_CANCER, names_to = "symptom", values_to = "value") %>%
  group_by(LUNG_CANCER, symptom) %>%
  summarise(prevalence = mean(value == 2)) %>%
  ggplot(aes(x = symptom, y = prevalence, fill = LUNG_CANCER)) +
  geom_col(position = "dodge", alpha = 0.8) +
  scale_fill_manual(values = c("NO" = "#2E9FDF", "YES" = "#E64B35")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Symptom Prevalence by Lung Cancer Status",
       x = "Symptom", y = "Prevalence", fill = "Lung Cancer")
```

The exploratory analysis reveals:

- The dataset is heavily imbalanced with **270 lung cancer cases (87.4%)** 
vs **39 non-cancer cases (12.6%)**
- **Age** shows overlapping distributions between groups, suggesting it 
may not be a strong predictor in this dataset
- All symptoms show higher prevalence in lung cancer patients, with 
**fatigue** and **smoking** showing the largest differences between groups

## Model Training

### Handling Class Imbalance

Given the significant class imbalance (87.4% YES vs 12.6% NO), we apply 
class weights to ensure the model pays adequate attention to the minority 
class (NO).

```{r train-model}
# Define class weights
class_weights <- c("NO" = 10, "YES" = 1)

# Train Random Forest model
set.seed(123)
rf_model <- randomForest(
  LUNG_CANCER ~ .,
  data = train_data,
  ntree = 500,
  mtry = 4,
  importance = TRUE,
  classwt = class_weights
)

# Print model summary
print(rf_model)
```

The Random Forest model was trained on 248 observations using 500 trees 
and 4 variables per split.

The OOB error rate of **10.08%** indicates good overall performance. 
The confusion matrix shows:

- **94.4% of lung cancer cases** correctly identified
- **59.4% of non-cancer cases** correctly identified

The lower sensitivity for non-cancer cases reflects the challenge of 
the class imbalance despite the use of class weights.

## Model Evaluation

We evaluate the model on the **test set** using standard classification 
metrics and the ROC curve to assess its discriminative ability.

```{r model-evaluation}
# Make predictions on test set
predictions <- predict(rf_model, test_data)
probabilities <- predict(rf_model, test_data, type = "prob")

# Confusion matrix
conf_matrix <- confusionMatrix(predictions, test_data$LUNG_CANCER,
                                positive = "YES")
print(conf_matrix)

# ROC curve
roc_obj <- roc(test_data$LUNG_CANCER, probabilities[, "YES"])

# Plot ROC curve
plot(roc_obj,
     main = paste("ROC Curve - AUC:", round(auc(roc_obj), 3)),
     col = "#E64B35",
     lwd = 2,
     print.auc = TRUE)
abline(a = 0, b = 1, lty = 2, col = "grey")
```

The model evaluation on the test set shows excellent performance:

- **AUC-ROC = 0.966:** outstanding discriminative ability, indicating 
the model can almost perfectly distinguish between lung cancer and 
non-cancer cases
- **Sensitivity = 98.1%:** the model correctly identifies 98.1% of 
lung cancer cases — critical for a screening tool where missing a 
positive case has serious consequences
- **Specificity = 14.3%:** the model struggles to identify non-cancer 
cases, reflecting the challenge of the class imbalance
- **Accuracy = 88.5%:** high overall accuracy, though largely driven 
by the majority class

The high sensitivity makes this model potentially valuable as a 
**preliminary screening tool**, where the priority is to identify 
all possible cancer cases rather than minimize false positives.

## Variable Importance

As in the previous Random Forest model, we examine which variables 
contribute most to predicting lung cancer status.

```{r variable-importance}
vip(rf_model,
    num_features = 15,
    aesthetics = list(fill = "#E64B35", color = "#E64B35", alpha = 0.8)) +
  theme_minimal() +
  labs(title = "Variable Importance - Lung Cancer Random Forest",
       subtitle = "Top 15 predictors of lung cancer",
       x = "Importance", y = "Variable")
```

The variable importance plot reveals interesting findings:

- **Alcohol consuming** is the strongest predictor, consistent with 
known associations between alcohol use and lung cancer risk
- **Peer pressure** ranks second, likely reflecting social influences 
on lifestyle risk factors such as smoking
- **Yellow fingers** ranks third, directly associated with heavy smoking
- **Age and Gender** show the lowest importance, suggesting that in 
this dataset symptoms and behavioral factors are stronger predictors 
than demographic variables

This contrasts with the diabetes model, where Age was the strongest 
predictor, highlighting how different diseases have different 
risk factor profiles.

## Conclusions

This analysis demonstrates the application of Random Forest machine 
learning to predict lung cancer status from clinical symptoms and 
lifestyle factors in a survey-based dataset.

**Key findings:**

1. The Random Forest model achieved an outstanding **AUC-ROC of 0.966**, 
indicating near-perfect discriminative ability between lung cancer 
and non-cancer cases.

2. The model showed very high **sensitivity of 98.1%**, correctly 
identifying almost all lung cancer cases — a critical property 
for a screening tool where missing a positive case has serious 
clinical consequences.

3. **Alcohol consuming, peer pressure and yellow fingers** were the 
strongest predictors of lung cancer, highlighting the importance 
of behavioral and lifestyle factors over demographic variables.

4. **Age and Gender** showed the lowest importance, contrasting with 
the diabetes model where age was the dominant predictor — demonstrating 
that different diseases have fundamentally different risk factor profiles.

5. Class imbalance (87.4% YES vs 12.6% NO) was addressed using class 
weights, though specificity remained low (14.3%).

**Limitations:**

- The dataset is small (309 patients) which limits the generalizability 
of the model
- Survey-based data may be subject to self-reporting bias
- The very high AUC may indicate overfitting due to the small dataset size
- External validation on an independent clinical dataset is needed

**Future directions:**

- Apply SMOTE to address class imbalance more effectively
- Validate the model on a larger, independent clinical dataset
- Compare performance with other algorithms such as XGBoost or 
Logistic Regression
- Incorporate additional clinical variables such as imaging data